Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-10-17 Paper/paper.tex
improved abstract

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the star's tidal distortion by
the companion black hole (BH) during inspiral, and (b) its potential tidal
disruption near merger.
%
In this paper, we present a Bayesian study of the measurability of
neutron star (NS) tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$
from GW observation(s) of disruptive NSBH coalescences, taking into account the
crucial effect of BH spins.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\simeq30$.
% 
Second, for (loud) signals with SNR $\gtrsim 23$ we can bound  $\lambdans$
by a factor of two.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median $\lambdans$ converges within $10\%$ of the true
value in approximately $20$ observations; (ii) in the same
number of observations, the associated $90\%$ credible intervals shrink to
$\pm 50\%$ of the true value.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

>>>>

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the tidal distortion of the
neutron star by its companion black hole during inspiral, and (b) its
potential tidal disruption near merger.
%
In this paper, we present a Bayesian study of the measurability of
neutron star tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$
using observation(s) of inspiral-merger GW signals from disruptive NSBH
coalescences, taking into account the crucial effect of black hole spins.
% 
First, we find that if non-tidal templates are used to estimate source 
parameters for an NSBH signal, the bias introduced in the estimation of
non-tidal physical parameters will only be significant for loud signals with
signal-to-noise ratios greater than $\simeq30$.
% 
For similarly loud signals, we also find that we can begin to put
interesting constraints on $\lambdans$ (factor of $1-2$) with individual
observations.
% 
Next, we study how a population of realistic NSBH detections will improve our
measurement of neutron star tidal deformability. For an astrophysically likely
population of {\it disruptive} NSBH coalescences, we find that $20-35$ events
are sufficient to constrain $\lambdans$ within $\pm 25-50\%$, depending on the
neutron star equation of state. For these calculations we assume that LIGO will
detect black holes with masses within the astrophysical {\it mass-gap}. In case
the mass-gap remains {\it preserved} in NSBHs detected by LIGO, we estimate that
approximately $25\%$ additional detections will furnish comparable $\lambdans$
measurement accuracy.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be planned for upcoming NSBH observations with the
LIGO-Virgo instruments.
\end{abstract}
